Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-25 paper2.qmd
section:code links; process:write,figures,math
section:all; process:write,figures,code

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement  [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Moreover, research on CJ's practical applicability highlights its versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. The study then addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's theory to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data. @sec-discussion discusses the implications of the findings and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

# Two prominent issues in CJ literature {#sec-theory-issues}

This section identifies and discusses the two prominent issues that can undermine the reliability and validity of CJ's trait estimates. @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

## The Case V and the statistical analysis of CJ data {#sec-theory-issue1}

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a trait scaling problem. Specifically, the model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons. In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge. In Case III, he posited that there was no correlation between stimuli. In Case IV, he assumed that the stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248–253].

@@

<!-- 3. The Case V and its simpler statistical representation: the BTL model -->
Notably, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli's discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). This substitution has little impact on the model's estimation or interpretation, as the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270]. Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. A thought experiment can illustrate this concept. In this experiment, the researcher observes the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ. Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area)

The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
:::

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Therefore, the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For instance, when researchers observe a sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of the discriminal difference then ensures reliable estimates for the texts’ discriminal processes [(citation needed?)]{style="color:red;"}.

<!-- 3. But does it hold?  -->
However, Thurstone argued that the assumption of equal dispersions may not be applicable when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, evidence of this violation may already be present in the CJ literature in the form of misfit statistics, which measure judgment discrepancies associated with specific stimuli [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For example, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. These discrepancies, in turn, suggest that the discriminal differences for "misfit" texts have broader distributions, indicating that their discriminal processes may also exhibit more variation than that of other texts. A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves. These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Thus, erroneously assuming equal dispersions between stimuli, can give rise to significant statistical and measurement issues. For instance, the model may overestimate the degree to which the outcome accurately reflects the "true" discriminal differences between stimuli. This overestimation can result in researchers drawing spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
The correlation, represented by the symbol $\rho$, measures how much a judge's perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. A similar thought experiment, as the one in @sec-theory-issue1a, can illustrate this concept. The experiment now assumes that the discriminal dispersions for both texts remain constant. @fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are highly correlated. Consequently, the discriminal difference is less likely to favor Text B over Text A (shaded gray area).

<!-- 2. But this is not the scenario we observe -->
Again, in experimental practice, researchers approach this process in reverse. They begin by observing the sample of outcomes favoring Text B over Text A and then use the BTL model to estimate the discriminal difference and the discriminal processes of the stimuli. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, then the model estimates a discriminal difference distribution that accurately reflects the "true" discriminal difference of the texts. This scenario is also illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts [(citation needed?)]{style="color:red;"}.

<!-- 2. But does it hold?  -->
Notably, Thurstone attributed the lack of correlation between stimuli to the cancellation of potential judges' biases. He argued that this cancellation resulted from two opposing and equally weighted effects occurring during pairwise comparisons [@Thurstone_1927b, pp. 268]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, it is easy to imagine at least two scenarios in which the zero correlation assumption is almost certainly invalid: when the pairwise comparison involves multidimensional, complex traits with heterogeneous stimuli and when an additional hierarchical structure is relevant to the stimuli.

<!-- 3. The first scenario where the assumption does not hold -->
In the first scenario, the intricate aspects of multidimensional, complex traits may introduce dependencies between the stimuli due to certain judges' biases that resist cancellation. Research on text quality suggests that when judges evaluate these traits, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases may ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp. 346]. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 4. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections created by additional hierarchical structures may further introduce dependencies between stimuli, creating a statistical phenomenon known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical handling of this additional source of dependence between stimuli has been inadequate. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) estimated BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. This approach, however, can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

@@

<!-- 5. But what are the challenges when this assumption is violated? -->
In any case, similar to @sec-theory-issue1a, assuming zero correlation between stimuli by neglecting additional relevant traits, excluding judges based on misfit statistics, or ignoring hierarchical (grouping) structures can cause significant statistical and measurement issues. In general, the model may over- or underestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. Such inaccuracies can result in spurious inferences about these differences and, by extension, about the stimuli's discriminal processes. This scenario is illustrated by @fig-correlation, when the model's discriminal difference distribution aligns with the thick continuous line for $\rho=0$, while the "true" discriminal difference follows any discontinuous line where $\rho \neq 0$.

In particular, neglecting relevant traits, such as judges' biases, can cause dimensional mismatches in the BTL model, artificially inflating the trait's reliability [@Hoyle_et_al_2023, pp. 341] or, worse, introducing bias into the trait's estimates [@Ackerman_1989]. Excluding judges based on misfit statistics risks discarding valuable information, which may further bias the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Finally, ignoring hierarchical structures may reduce the precision of model parameter estimates, potentially overestimating the trait's reliability [@Hoyle_et_al_2023, pp. 482].

@@

<!-- 4. segway for next section -->
In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of the trait estimates. This overreliance may also undermine their validity [@Perron_et_al_2015, pp. 2], especially when combined with the disconnect between CJ's approach to trait measurement and hypothesis testing. However, the structural approach to causal inference can address these issues by providing a systematic and integrated framework that enhances statistical accuracy while strengthening measurement reliability and validity.

@@

# Extending Thurstone's theory {#sec-theoretical}

@@

@@

>>>>

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on the validity of CJ scores indicates their capacity to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Moreover, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues that can undermine the reliability and validity of CJ's trait estimates. The first issue arises from the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second stems from the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. This study addresses these issues by extending Thurstone's general form through a systematic and integrated approach based on Causal and Bayesian inference methods.

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model for the analysis of dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's key insights.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, since the individual discriminal processes of the stimuli are not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus with a discriminal process positioned further along the trait continuum as possessing more of the trait [@Bramley_2008, pp. 251]. This suggests that the relative distance between stimuli, rather than their absolute positions on the continuum, likely defines the outcome of pairwise comparisons. Indeed, the theory assumes that the difference between the underlying discriminal processes of the stimuli, referred to as the *discriminal difference*, determines the observed dichotomous outcome. Furthermore, the theory assumes that because the individual discriminal processes form a Normal distribution on the continuum, the discriminal difference will also conform to a Normal distribution [@Andrich_1978, pp. 452]. In this distribution, the mode (mean) represents the relative separation between the stimuli, and its dispersion indicates the variability of that separation.

@@

# Two prominent issues in CJ practice {#sec-theory-issues}

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that his general form, described in @sec-thurstone_theory, led to a *trait scaling problem*. Specifically, the model required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b, pp. 267]. For instance, in a CJ experiment with five texts, the general form would require estimating $20$ parameters: five modal discriminal processes, five discriminal dispersions, and $10$ correlations --one per comparison (see @tbl-thurstone_cases). However, a single judge could provide only ${5 \choose 2} = 10$ unique comparisons, an insufficient data set to estimate the required parameters.

<!-- 2. Thurstone's Cases and their simplifying assumptions -->
To address this issue and facilitate the practical implementation of the theory, Thurstone developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model. In Case I, Thurstone postulated that pairs of stimuli would maintain a constant correlation across all comparisons. In Case II, he allowed multiple judges to undertake comparisons instead of confining evaluations to a single judge. In Case III, he posited that there was no correlation between stimuli. In Case IV, he assumed that the stimuli exhibited similar dispersions. Finally, in Case V, he replaced this assumption with the condition that stimuli had equal discriminal dispersions. @tbl-thurstone_cases summarizes the assumptions of the general form and the five cases. For a detailed discussion of these cases and their progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248–253].

@@

<!-- 3. But are our specific problems? -->
It is evident that although Thurstone developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269] and cautioned that its use "should not be made without experimental test" [@Thurstone_1927b, pp. 270], the CJ literature has nevertheless relied the most on it in practice. This preference is particularly notable given that Case V incorporates the most extensive set of simplifying assumptions among all the cases [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677] (see @tbl-thurstone_cases).

Furthermore, since Thurstone's primary goal was to define a psychological scale and "allocate the compared stimuli on this continuum" [@Thurstone_1927b, pp. 269], he did not provide guidance on how to use these measurements when, for instance, researchers aim to test hypotheses related to the measured traits. Although the CJ practice has circumvented this issue by using scores' point estimates or their transformations for such analyses, a fundamental question remains: Is this method appropriate for addressing these inquiries?

<!-- 4. Section division -->
Thus, this section discusses these two prominent issues, as they could undermine the reliability and validity of CJ's trait estimates [@Perron_et_al_2015, pp. 2]. Specifically, @sec-theory-issue1 examines the the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. Moreover, @sec-theory-issue2 focuses on the apparent disconnect between the approaches to trait measurement and hypothesis testing in CJ.

## The overreliance on Case V in the statistical analysis of CJ data {#sec-theory-issue1}

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
As previously discussed, despite relying on the most extensive set of simplifying assumptions [@Bramley_2008, pp. 253; @Kelly_et_al_2022, pp. 677], Case V remains the most widely used case in the CJ literature. This popularity stems mainly from its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors the assumptions of Case V, with one notable distinction: whereas Case V assumes a Normal distribution for the stimuli discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). As noted by @vanderLinden_et_al_2017_I [pp. 16], this substitution has little impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation. Furthermore, this minimal impact is reinforced by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 2. But what are limitations of Case V (and the BTL model)? -->
However, as mentioned earlier, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270]. Furthermore, he acknowledged that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice, namely the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- 1. But does it hold?  -->
Thurstone argued that the assumption of equal dispersions may not apply when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. Indeed, the CJ literature may already provide evidence of this violation through "misfit" statistics. For example, labeling texts as *misfits* indicates that comparisons involving these texts result in more judgment discrepancies than those involving other texts [@Pollitt_2004, pp. 11; @Pollitt_2012b, pp. 290]. These discrepancies imply that the discriminal differences for *misfit* texts have broader distributions, further suggesting that their discriminal processes exhibit more dispersion than those of other texts. This idea can be illustrated using @fig-discriminal_process, where text B shows a wider discriminal dispersion than text A. As a result, the texts' discriminal processes overlap more, making it difficult for a judge to determine which text possesses more of the trait.

A similar line of reasoning applies to the concept of "misfit" judges, whose evaluations deviate substantially from the shared consensus due to the unique characteristics of the stimuli or the judges themselves [@Pollitt_2012a, pp. 164-165; @Pollitt_2012b, pp. 289-290; @vanDaal_et_al_2016, pp. 4; @Goossens_et_al_2018, pp. 20]. These "misfit" judges and their associated deviations can give rise to additional statistical and measurement issues, which we discuss in more detail in @sec-theory-issue1b.

@@

<!-- 2. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. A thought experiment can illustrate this idea. In this experiment, researchers can observe the discriminal processes for the texts depicted in @fig-discriminal_process. Furthermore, the experiment assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. @fig-dispersion reveals that an increase in the uncertainty associated with the perception of Text B compared to Text A $(\sigma_{B} - \sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B>A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B} - \sigma_{A}=0)$, the discriminal difference is more narrow compared to situations where their dispersions differ. Consequently, the discriminal difference is more likely to favor Text B over Text A (shaded gray area).

The effect of dispersion discrepancies and stimuli correlation on the distribution of the discriminal difference.
:::

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, the thought experiment occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. Consequently, the ability of the outcome to reflect *true* differences between stimuli largely depends on the validity of the model's assumptions [@Kohler_et_al_2019, pp. 150], particularly the assumption of equal dispersions. For example, when researchers correctly assume equal dispersions between texts, the BTL model estimates a discriminal difference distribution that accurately represents the *true* discriminal difference between them. This scenario is illustrated in @fig-dispersion, when the model's discriminal difference distribution aligns with the *true* discriminal difference distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this discriminal difference then ensures reliable estimates for the texts’ discriminal processes.

<!-- 2. So, what happens when the assumption does not hold? -->
However, erroneously assuming equal dispersions between stimuli can give rise to significant statistical and measurement issues. Specifically, the model may overestimate the trait's reliability, that is, the degree to which the outcome accurately reflects the *true* discriminal differences between stimuli. This overestimation, in turn, results in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, about the underlying discriminal processes of stimuli. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the *true* discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Additionally, if researchers recognize that misfit statistics highlight these critical differences in dispersions, the conventional CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

<!-- 1. But does it hold?  -->
In contrast, Thurstone assumed that stimuli were uncorrelated because judges' biases, arising from two opposing and equally weighted effects occurring during the pairwise comparisons, canceled each other out [@Thurstone_1927b, pp. 268]. @Andrich_1978 provided a mathematical demonstration of this cancellation using the BTL model under the assumption of discriminal processes with additive biases. However, the assumption of zero correlation is not met in practice in at least two scenarios. First, when intricate aspects of multidimensional, complex traits or heterogeneous stimuli influence judges' perceptions, generating biases that resist cancellation. Second, when an additional hierarchical structure is relevant to the stimuli.

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases in their judgments that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could ultimately introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II, pp. 346]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 3. The second scenario where the assumption does not hold -->
In the second scenario, the shared context or inherent connections introduced by additional hierarchical structures may create dependencies between stimuli, a statistical phenomenon known as clustering [@Everitt_et_al_2010]. Although the CJ literature acknowledges the existence of such hierarchical structures, the statistical approaches to account for this additional source of dependence have been insufficient. For instance, when CJ data incorporates multiple samples of stimuli from the same individuals, researchers frequently rely on (averaged) point estimates of the BTL scores to conduct subsequent analyses and tests at the individual hierarchical level [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021]. However, this approach can introduce additional statistical and measurement issues, which we discuss in greater detail in @sec-theory-issue2.

@@

<!-- 4. What does the zero correlation between stimuli imply? -->
According to the theory, the correlation $\rho$ measures how much the judges' perception of a specific trait in one stimulus depends on their perception of the same trait in another. As with the discriminal dispersions, this correlation shapes the distribution of the discriminal difference, directly impacting the outcomes of pairwise comparisons. A similar thought experiment, as the one depicted in @sec-theory-issue1a, can illustrate this idea. The experiment now assumes that the discriminal dispersions for both texts remain constant. @fig-correlation reveals that as the correlation between the texts increases, the distribution of their discriminal difference becomes narrower. This narrowing affects the area under the curve where the discriminal difference distinctly favors Text B over Text A, denoted as $P(B > A)$, thus influencing the comparison outcome. Furthermore, the figure shows that when two texts are independent or uncorrelated $(\rho=0)$, their discriminal difference is less narrow compared to scenarios where the texts are highly correlated. Consequently, the discriminal difference is less likely to favor Text B over Text A (shaded gray area).

<!-- 5. But this is not the scenario we observe -->
Again, in experimental practice, researchers approach this process in reverse. They begin by observing the outcomes and use the BTL model to estimate the stimuli' discriminal difference and discriminal processes. Given that the BTL model assumes independent discriminal processes across comparisons, if this assumption holds, the model estimates a discriminal difference distribution that accurately reflects the *true* discriminal difference between the texts. This scenario is illustrated in @fig-correlation when the discriminal difference distribution of the model aligns with the *true* distribution, represented by the thick continuous line corresponding to $\rho=0$. Once more, the accuracy of the discriminal difference ensures reliable estimates for the discriminal processes of the texts.

<!-- 6. But what are the challenges when this assumption is violated? -->
However, erroneously assuming zero correlation between stimuli can lead to significant statistical and measurement issues. Specifically, neglecting judges' biases or hierarchical structures can create dimensional mismatches in the model, leading to the over- or underestimation of trait reliability [@Hoyle_et_al_2023, pp. 341, 482]. In other words, the outcome may fail to accurately reflect the *true* discriminal differences between stimuli. These inaccuracies can result in spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, the underlying discriminal processes  [@Ackerman_1989]. This issue is illustrated in @fig-correlation when the model discriminal difference distribution follows the thick continuous line $(\rho = 0)$, while the *true* discriminal difference follows any discontinuous line where $(\rho \neq 0)$. Conversely, excluding judges based on misfit statistics risks discarding valuable information, potentially biasing the trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. Again, the direction and magnitude of these biases are often unpredictable, as they depend on which judges are excluded from the analysis.

@@

<!-- 4. segway for next section -->
In aggregate, excessive reliance on Thurstone's Case V assumptions in the statistical analysis of comparative data can compromise the reliability of trait estimates. This overreliance may also undermine their validity [@Perron_et_al_2015, pp. 2], particularly when coupled with the disconnect between CJ's approach to trait measurement and hypothesis testing. However, the structural approach to causal inference can address these issues by offering a systematic and integrated framework that strengthens measurement reliability and validity while enhancing the statistical accuracy of hypothesis tests.

@@

# Extending Thurstone's general form {#sec-theoretical}

@@

Notice,

@@

Sample-comparison model, including sample and comparison mechanisms and assuming equal discriminal dispersions for the student's traits
:::

::: {#fig-cj16 layout-ncol=2}
